---
title: "descriptive"
format:
  html:
    theme: cosmo
    toc: true
    code-fold: true
---

## 3. Descriptive Statistics

### 3.1 Overview

This section summarizes the key descriptive statistics of the NHL skater performance dataset.
We compute central tendencies, variability, distribution patterns, and identify any outliers that may influence later modeling steps.

```{r}
library(dplyr)
library(ggplot2)
library(knitr)

# Load data
skaters <- read.csv("D:/GMU homeworks/STAT515/final proj/skaters.csv")

# Convert numeric-looking character columns
skaters_clean <- skaters %>%
  mutate(across(where(is.character), ~ suppressWarnings(as.numeric(.))))

# Summary statistics for numeric variables
numeric_summary <- skaters_clean %>%
  select(where(is.numeric)) %>%
  summarise(across(everything(),
                   list(
                     mean = ~ mean(.x, na.rm = TRUE),
                     median = ~ median(.x, na.rm = TRUE),
                     sd = ~ sd(.x, na.rm = TRUE),
                     min = ~ min(.x, na.rm = TRUE),
                     max = ~ max(.x, na.rm = TRUE)
                   ),
                   .names = "{.col}_{.fn}"
  ))

kable(numeric_summary, caption = "Descriptive Statistics for Numeric Variables")
```

------------------------------------------------------------------------

### 3.2 Distribution of Key Variables

Below we visualize distributions for important performance variables such as shifts, TOI, xGoalsForAfterShifts, and xGoalsAgainstAfterShifts.

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)

# Desired variables
vars_to_plot <- c(
  "shifts",
  "toi",
  "xGoalsForAfterShifts",
  "xGoalsAgainstAfterShifts"
)

# Keep only columns that exist
vars_present <- intersect(vars_to_plot, names(skaters_clean))

if (length(vars_present) == 0) {
  stop("None of the requested variables exist in skaters_clean.")
}

df_long <- skaters_clean %>%
  pivot_longer(
    cols = all_of(vars_present),
    names_to = "variable",
    values_to = "value"
  )

ggplot(df_long, aes(x = value)) +
  geom_histogram(
    bins = 40,
    fill = "steelblue",
    color = "white",
    alpha = 0.7
  ) +
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  labs(
    title = "Distribution of Key Performance Metrics",
    x = "Value",
    y = "Frequency"
  ) +
  theme_minimal()

```

------------------------------------------------------------------------

### 3.3 Detecting Outliers

Outliers can distort averages and influence statistical modeling.\
We detect numeric outliers using the 1.5 × IQR rule.

```{r}
outlier_detection <- function(x) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  sum(x < (q1 - 1.5 * iqr) | x > (q3 + 1.5 * iqr), na.rm = TRUE)
}

outliers <- skaters_clean %>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), outlier_detection))

kable(outliers, caption = "Number of Outliers Detected per Variable")
```

------------------------------------------------------------------------

### 3.4 Correlation Analysis

We examine relationships between selected numeric variables to identify potential multicollinearity and performance-related patterns.

```{r}
library(corrplot)

corr_data <- skaters_clean %>%
  select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs")

corrplot(corr_data, method = "color", type = "upper", tl.col = "red")
```

------------------------------------------------------------------------

### 3.5 Summary

-   The dataset contains a wide range of performance metrics with varying spreads.
-   Some variables show skewness, requiring possible transformation.
-   Several variables contain notable outliers.
-   Correlation analysis suggests meaningful relationships that will guide predictive modeling.

This descriptive statistics section provides a foundation for deeper inferential analysis and aligns with the quantitative structure of academic data analytics research.
